An investigation into the contentious phylogenetic results of the *Mixta* species

### 2021-05-04 Manuscript/Manuscript_Paper.Rmd
Update Manuscript_Paper.Rmd

Kimberly N. Hinz, John Stavrinides^1^

@@

# Discussion

# Acknowledgements

>>>>

Kim N. Hinz, John Stavrinides^1^

@@

# Discussion
Phylogenetic software programs vary on which models and parameters they have available, and these may vary within a program depending on the chosen method. MEGAX [@Kumar2018], for example, offers a wide range of models and additional parameters for the building of phylogenetic trees; however, the options are reduced for genetic distance calculation. A vital practise of selecting the most accurate model for your data is by conducting model testing. Model testing provides a list of models with additional parameters in the order of which model explains the most variation in the sequences’ differences. Selecting the most accurate model is always recommended; however, it is not always possible. My results show that slight differences in the taxa’s relationships occur when the more precise model is not used. For some genes, the phylogenetic signal of which taxon was the closest relative to *Mixta* was strong enough despite inaccuracies in the model. Nonetheless, there is a risk that the wrong species is labelled as the closest relative, and this poses problems if that species belongs to a separate genus or order. The species of interest may be wrongfully defined as belonging to a specific genus or order, or the genes may be inappropriately labelled as results from horizontal gene transfer (HGT).

Additionally, the chosen method may contribute to contentious results. In order to use the most accurate model, I had to first run a phylogenetic analysis and then extract the genetic distance matrices. For the best available model for genetic distance, I ran the genetic distance analysis directly and thus, forwent the phylogenetic tree step. Even when the most accurate and the best available models were the same, the results for the closest relative were different. This suggests that the two methods calculate taxa relatedness differently despite using the same models and parameters. One explanation is the Gamma parameter (+G). The genetic distance method uses a continuous distribution in which 1.00 is the default, while the phylogenetic tree method uses a discrete Gamma parameter set to 5 categories. Therefore, diligence must be given when selecting software programs, methods, and models.

Phylogenetic analysis is an accepted method for understanding the evolutionary relatedness [@Boundless2021] while site identity is used to confirm identification and taxon boundaries [@Ciufo2018]. These two methods were conducted in this study as both nucleotide (NTS) and amino acid (AAS) sequences. *P. septica* was the closest relative to the majority of *Mixta* homologues in all four analyses. This lends strong support that *Mixta* and *P. septica* share a recent common ancestor, followed by other *Pantoea* species, *Erwinia*, and then *Tatumella*. This agrees with *Mixta* being previously known as part *Pantoea*. Additionally, there were no *Mixta* sequences that were closely related to *P. syringae* which demonstrates that this species is a suitable outgroup.

For both methods, fewer genes were labelled as being closely related to *P. septica* when using the AAS versus the NTS. This difference may be to a loss of phylogenetic signal when using amino acid sequences caused by codon degeneracy. Nevertheless, sequence type, as well as different evolutionary models between NTS and AAS, affects the results. AAS may be better suited for uncovering taxonomic relationships at the genus or higher level and NTS may be more appropriate for the species level.

When considering the spatial effects of the closest relatives, the uniform pattern of the genes related to *P. septica* further lends support that this species and *Mixta* share a recent common ancestor. Additionally, the other taxa followed the same pattern. If the genes that were closely related to non-*Pantoea* species are a result of HGT, then this suggests that HGT may be a common occurrence within the *Erwiniaceae*. A second explanation for the even distribution patterns in the closest relatives is that the genes are highly conserved. Two or more species may have been tied for closest relative and the R program may have selected one based another characteristic, such as alphabetical order. Nevertheless, further research into these genes and the frequency of HGT events with *Mixta* and other genera within the *Erwiniaceae* is needed.
